Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2016-03-20 RMS/submission.tex
modified fontsize in verbatim, shorten the paper within 8 pages.
edited the appendix

\usepackage{alltt}
%\usepackage{epic}
%\usepackage{pstricks}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\begin{figure}[!t]
\hrule\vspace{1mm}
  \begin{algorithmic}[1]
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts} \label{l:1stline}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return; \label{l:return}
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to execute the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} \Comment{the previous job finishes} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
  \State \Call{$reportTaskError$}{$p$}; \Comment{the task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\hrule
  \caption{The C-Like Pseudo-code of $schedule()$}
  \label{a:schedule}
\end{figure}

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, of a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
 op _____ : TaskList Timer 
            SysTasks Hardware Object 
              ~> System [ctor] .
 mb (L T STS HW < O:Oid : IntSrc |>) 
      : System .
\end{verbatim}

@@

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
Timed behaviors of the system consist of two parts: the execution of
tasks and the execution of the interrupt source. Both are modeled
together by the following single ``standard'' tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
 crl [tick]:
   {SYSTEM} 
   => {delta(SYSTEM, R)} in time R 
   if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable \verb|R| is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

@@

Another objective is to verify the correctness of the implementation.
The atomic proposition \verb|correct| is hence defined to hold if the
running periodic task is the one requested to be executed with the
highest priority:
\begin{verbatim}
 op correct : -> Prop [ctor] .
 ceq {L T STS HW ISRC} |= correct
     = if ID :: MaybeNat 
       then shouldRun(ID, L)
       else true fi
     if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is \verb|READY|,
\verb|RUNNING| or \verb|INTERRUPT|. Note that in this paper, we do not
care the behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
 (mc init |=u (correct U taskTimeout) 
              \/ ([]correct) .)
\end{verbatim}

@@

@@

  % legend
  \begin{scope}[xshift = -1.2cm]  
    \draw [fill=gray!25,draw=gray!50,dashed] (1.5,4) rectangle +(8,1);
    \draw (1.5,4.5) node [above right] {LEGEND};
    \draw [init] (2,4.4) -- +(0,-0.1);
    \draw (2.2,4.55) node [below right] {: Initiation of Tasks};
    \draw [complete] (6,4.2) -- +(0,0.1);
    \draw (6.2,4.55) node [below right] {: Completion of Tasks};
  \end{scope}

\end{tikzpicture}
\caption{A counterexample of schedulability in Scenario~(iv). The first
  job instance of $\tau_3$ misses its deadline at time $15ms$, which
  is the initiation time for the second job instance of $\tau_3$.}
\label{f:counterexample}
\end{figure*}

@@

Therefore, we achieve the following theorem, showing that the results
in Section~\ref{ss:results} are complete.
\begin{theorem} 
Our approach using untimed model checking to verify the schedulability
and the correctness of our model is complete.
\end{theorem}
\begin{IEEEproof}
By showing that our model is time-robust and that the two defined
atomic propositions--\verb|taskTimeout| and \verb|correct|--are
tick-stabilizing, then using Theorem~\ref{t:completeness}.
\end{IEEEproof}

@@

>>>>

\usepackage{alltt}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepackage{mathabx}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}

% Define the fontsize in environment {verbatim}
\makeatletter
\def\verbatim{\small\@verbatim \frenchspacing\@vobeyspaces \@xverbatim}
%\def\verbatim@font{\small\ttfamily}
\makeatother

@@

\begin{figure}[!t]
\hrule\vspace{1mm}
  \begin{algorithmic}[1]\small
\Function{$schedule$}{$ $}{}
  \State \Call{$int\_o\!f\!\!f$}{$ $}; \Comment{to disable interrupts} \label{l:1stline}
  \State \Call{$updateStatus$}{$taskList$}; \label{l:updatestatus}
  \State $timer = timer + 1$; \label{l:timer} \label{l:inc}
  \State $p = taskList$;
  \While{$p$} \label{l:startrun1st}
    \If{$p\rightarrow status == \textit{INTERRUPT}$}
      \State \Return; \label{l:return}
    \ElsIf{$p\rightarrow status == \textit{READY}$}      
      \State $p\rightarrow status = \textit{RUNNING}$;
      \State \Call{$int\_on$}{$ $}; \Comment{to enable interrupts} \label{l:endrun1st}
      \State $p\rightarrow function()$; \Comment{to execute the task} \label{l:function}
      \State \Call{$int\_o\!f\!\!f$}{$ $};
      \State $p\rightarrow status = \textit{DORMANT}$;
    \EndIf
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
\Function{$updateStatus$}{$p$}
  \While{$p$}
    \If{$p\rightarrow status == \textit{RUNNING}$} \label{l:startupdate}
      \State $p\rightarrow status = \textit{INTERRUPT}$;
    \EndIf
    \If{$timer~\%~(p\rightarrow period) == 0$} \Comment{the task should be initiated}
      \If{$p\rightarrow status == \textit{DORMANT}$} \Comment{the previous job finishes} 
        \State $p\rightarrow status = \textit{READY}$;
      \Else \Comment{the status is \textit{READY} or \textit{INTERRUPT}}
  \State \Call{$reportTaskError$}{$p$}; \Comment{task misses its deadline}
      \EndIf
    \EndIf \label{l:endupdate}
    \State $p = p\rightarrow next$;
  \EndWhile
\EndFunction
  \end{algorithmic}
\hrule
  \caption{The C-Like Pseudo-code of $schedule()$}
  \label{a:schedule}
\end{figure}

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, of a sort \verb|System|\footnote{Some variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : TaskList Timer 
             SysTasks Hardware Object 
               ~> System [ctor] .
  mb (L T STS HW < O : IntSrc |>) : System .
\end{verbatim}

@@

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
Timed behaviors of the system consist of two parts: the execution of
tasks and the execution of the interrupt source. Both are modeled
together by the following single \emph{standard} tick
rule~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable \verb|R| is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

@@

Another objective is to verify the correctness of the implementation.
The atomic proposition \verb|correct| is hence defined to hold if the
running periodic task is the one requested to be executed with the
highest priority:
\begin{verbatim}
  op correct : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= correct
      = if ID :: MaybeNat then shouldRun(ID, L)
        else true fi
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is \verb|READY|,
\verb|RUNNING| or \verb|INTERRUPT|. Note that in this paper, we do not
care the behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
  (mc init |=u ([]correct) 
               \/ (correct U taskTimeout) .)
\end{verbatim}

@@

  % legend
  \begin{scope}[xshift = -1.2cm]  
    \draw [fill=gray!25,draw=gray!50,dashed] (1.5,4) rectangle +(8,1);
    \draw (1.5,4.5) node [above right] {LEGEND};
    \draw [init] (2,4.4) -- +(0,-0.1);
    \draw (2.2,4.55) node [below right] {: Initiation of Tasks};
    \draw [complete] (6,4.2) -- +(0,0.1);
    \draw (6.2,4.55) node [below right] {: Completion of Tasks};
  \end{scope}
\end{tikzpicture}
\caption{A counterexample of schedulability in Scenario~(iv). The first
  job instance of $\tau_3$ misses its deadline at time $15ms$, which
  is the initiation time for the second job instance of $\tau_3$.}
\label{f:counterexample}
\end{figure*}

@@

@@

Therefore, we achieve the following theorem, showing that the results
in Section~\ref{ss:results} are complete.
\begin{theorem}
\label{t:main}
Our approach using untimed model checking to verify the schedulability
and the correctness of our model is complete.
\end{theorem}
\begin{IEEEproof}
By showing that our model is time-robust and that the two defined
atomic propositions--\verb|taskTimeout| and \verb|correct|--are
tick-stabilizing, then using Theorem~\ref{t:completeness}.
\end{IEEEproof}

@@

\appendix[Proof of Theorem~\ref{t:main}]
\setcounter{theorem}{0}
\renewcommand{\thetheorem}{\thesection.\arabic{theorem}}
\newcommand{\lrps}[2]{\rightarrow^{#1}_{#2}}

We show here a detailed proof of Theorem~\ref{t:main}.

First we need to introduce some more preliminaries
from~\cite{DBLP:journals/entcs/OlveczkyM07a}. A term $t$ is called
\emph{ground} if it contains no variables. For a set $P\subseteq AP$
of atomic propositions and ground terms $t,t'$, we write $t\simeq_P
t'$ (or $t\simeq t'$ for simplicity when $P$ is implicit) if $t$ and
$t'$ satisfy exactly the same set of propositions from $P$.

The following lemma is needed to prove the time-robustness of our
model:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{l:timerobustness}
Let $\cal R$ be an object-oriented specification with a standard tick
rule, let the infinity element \verb|INF| be the only element in the
time domain which is not a normal time value, and let the time domain
be \emph{linear}. Then $\cal R$ is time-robust if the following
conditions are satisfied for all appropriate ground terms $t$ and
$r,r'$:

\noindent
(i) \verb|mte(delta(|$t$\verb|,|$r$\verb|))| $=$
\verb|mte(|$t$\verb|) monus |$r$, for all $r\le$
\verb|mte(|$t$\verb|)|;

\noindent
(ii) \verb|delta(|$t$\verb|,|$0$\verb|)| $= t$;

\noindent
(iii) \verb|delta(delta(|$t$\verb|,|$r$\verb|),|$r'$\verb|)| $=$
\verb|delta(|$t$\verb|,|$r+r'$\verb|)|, for $r+r'\le$
\verb|mte(|$t$\verb|)|;

\noindent
(iv) \verb|mte(|$\sigma(l)$\verb|)|$= 0$ for each ground instance
$\sigma(l)$ of each left-hand side of an instantaneous rewrite rule.
\end{lemma}

Each one-step rewrite can be categorized and then tick-invariance can
be defined:
\begin{definition}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
A one-step rewrite $t\lrps{r}{1}t'$ using a tick rule and having
duration $r$ is:\\
- a \emph{maximal tick step}, written $t\lrps{r}{max}t'$, if there is
no time value $r'>r$ such that $t\lrps{r'}{1}t''$ for some $t''$; \\
- an \emph{$\infty$ tick step}, written $t\lrps{r}{\infty}t'$, if
for each time value $r'>0$, there is a tick rewrite step
$t\lrps{r'}{1}t''$; and \\
- a \emph{non-maximal tick step} if there is a maximal tick step
$t\lrps{r'}{max}t''$ for $r'>r$.
\end{definition}
\begin{definition}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
  A time-robust specification $\cal R$ is \emph{tick-invariant} with
  respect to a set $P$ of propositions if and only if it is the case
  that $t\simeq_P t'$ holds for each non-maximal or $\infty$ tick step
  $t\lrps{r}{}t'$.
\end{definition}

The following lemma is needed to prove the tick-invariance of our
defined propositions:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
Let $\cal R$ be a time-robust object-oriented specification with a
standard tick rule, let the infinity element \verb|INF| be the only
element in the time domain which is not a normal time value, and let
the time domain be \emph{linear}. Then $\cal R$ is tick-invariant with
respect to a set $P$ of atomic propositions if \verb|{|$t$\verb|}|
$\simeq_P$ \verb|{delta(|$t$\verb|,|$r$\verb|)}| for all $t,r$ with
$r<$ \verb|mte(|$t$\verb|)|.
\end{lemma}

Now we can present the detailed proof of Theorem~\ref{t:main}:
\noindent
\begin{IEEEproof}[Proof of Theorem~\ref{t:main}]
first
\end{IEEEproof}
